---
title: "Loan_EDLD629_HW1"
author: "Christopher Loan"
date: "4/9/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Prep-work

## Packages & Data

```{r}
library(tidyverse)
library(lme4)
library(performance)
```

```{r}
dat <- 
  rio::import(here::here(
    'Data',
    'longitudinal-sim.csv'
  )) %>% 
  mutate(sid = as.character(sid))
```

## Pivot & Make Wave Variable

```{r}
temp_dat <-
  dat %>% 
  pivot_longer(starts_with('g'), 
               names_to = 'time', 
               values_to = 'scores') 

key <- tibble(
  time = unique(temp_dat$time), 
  wave = 1:length(unique(temp_dat$time))-1
                  ) 

long_dat <- 
  left_join(temp_dat, key) %>% 
  mutate(grade = factor(parse_number(time)))

head(long_dat,1)
```

# Part A

## Model 1. Unconditional growth model with random intercepts and parallel slopes

```{r}
mod1 <- 
  long_dat %>% 
  lmer(formula = 
         scores ~ grade + (1 | sid), 
       control = lmerControl(optimizer = "bobyqa")
         )
```

## Model 2. Conditional growth model with random intercepts, parallel slopes, and grade-level fixed effects

```{r}
mod2 <- 
  long_dat %>% 
  lmer(formula = 
         scores ~ grade + wave + (1 | sid), 
       control = lmerControl(optimizer = "bobyqa")
         )
```

## Model 3. Unconditional growth model with random intercepts and random slopes

```{r}
mod3 <- 
  long_dat %>% 
  lmer(formula = 
         scores ~ wave + (wave | sid), 
       control = lmerControl(optimizer = "bobyqa")
         )
```

## Model 4. Conditional growth model with random intercepts, random slopes, and grade-level fixed effects

```{r}
mod4 <- 
  long_dat %>% 
  lmer(formula = 
         scores ~ grade + wave + (wave | sid), 
       control = lmerControl(optimizer = "bobyqa")
         )
```

# Part B

## Model 1 vs 2
```{r}
anova(mod1, mod2)
compare_performance(mod1, mod2)
```

## Model 2 vs 3

```{r}
anova(mod2, mod3)
compare_performance(mod2, mod3)
```

## Model 3 vs 4

```{r}
anova(mod3, mod4)
compare_performance(mod3, mod4)
```

## Which model fits the data best?

Four linear growth models were estimated to assess student scores on a seasonally administered test:

* Model 1: Unconditional growth model with random intercepts and parallel slopes
* Model 2: Conditional growth model with random intercepts, parallel slopes, and grade-level fixed effects
* Model 3: Unconditional growth model with random intercepts and random slopes
* Model 4: Conditional growth model with random intercepts, random slopes, and grade-level fixed effects

Models were compared with successive likelihood ratio tests (LRTs), AIC, and BIC. Significant LRTs, lower AICs, and lower BICs all suggested that successive (more complex) models continued to fit the data better until model 4 was identified as the optimal model.

# Part C

```{r}
summary(mod4)
confint(mod4)
```

          2.5 %      97.5 %
.sig01        9.986782  10.1886483
.sig02        0.316260   0.3457314
.sig03        1.144692   1.1720972
.sigma        4.697507   4.7304326
(Intercept) 188.652397 188.9300296
grade4       -8.258682  -8.0773529
grade5      -16.850919 -16.5328461
wave          6.147690   6.2063883

A multilevel model was fit to n = 22,500 students' data to see growth in a seasonally-administered benchmark assessment over 3 years. A model with random intercepts, random slopes, and grade-level fixed effects was found to fit the data best (see above). 

The average score on the assessment was 188.79 (95% CI = [188.65 ,188.93]), with a standard deviation of 10.09 (95% CI = [9.99, 10.12]). There was an average growth of 6.18 (95% CI = [6.15, 6.20]), with a standard deviation of 1.16 (95% CI = [1.14, 1.17]). Random effects had a medium correlation (0.33; 95% CI = [0.32, 0.35]), meaning those with higher baseline scores grew more rapidly, while those with lower baseline scores grew slower. 

Fixed effects comparison of grade 4 (estimate = -8.17, 95% CI [-8.26,  -8.08]) and grade 5 (estimate = -16.69, 95% CI [-16.85, -16.53]) to grade 3 suggest a decrease in score over time. However, when the effect of this is tempered with the increase in wave over time, the total observed effect is still growth. 

For example, if we look at grade 4, we see a drop by 8.17 estimated for the comparison to grade 3, but have approximately 6.18 points increase per wave over the past 3 waves (so 6.18*3 = +18.54), which corresponds to an average change of 10.37.

Small, positively correlated fixed effects of the intercept with grade 4 (0.05) and grade 5 (0.12), suggest those with higher intercepts have positive (small) differences in grade4 and grade5 compared to grade3. Large, negative fixed effect correlation of wave with grade4 (-0.71) and grade5 (-0.81) suggest those with higher slopes demonstrated lower estimates at grade 4 and grade 5, compared to grade 3. 

Taken together, this model suggests a fan-closed type of growth, with participants who show more growth demonstrated larger degrees of loss over the summer (i.e., greater magnitude of "summer slump"), but with participants converging in scores overall.

# Part 3: Plots of the model fit

```{r}
plot_dat <- 
  long_dat %>% 
  filter(sid %in% c('0001-01-01', 
                    '0001-01-02', 
                    '0001-01-03'))

plot_dat <- plot_dat %>% 
  mutate(prediction = 
           predict(mod4, newdata = plot_dat)
  )
```

```{r}
plot_dat %>% 
  ggplot(aes(group = sid)) +
  geom_point(
    aes(x = wave, y = scores)) + 
  geom_line(
    aes(x = wave, y = prediction), 
    color = '#67E0CF', 
    alpha = 0.7,
    size = 1.5) +
  facet_wrap(~sid) + 
  theme_minimal() +
  labs(title = 'Student scores from grade 3 to 5',
  caption = 'points = observed values
  lines = model predicted values')
```

